---
title: "Lecture 1"
author: "Han Zhang"
date: "2/1/2021"
output: pdf_document
---

Population is 10000 units following Beta distribution.

```{r}
population <- rbeta (10000, 1, 2)
plot (density (population))
# population mean
mean(population)
# population variance
var(population)

```

Create samples from the population with size 1000


```{r}
sample = sample(population, size = 1000)
plot (density (sample))
# sample mean
mean(sample)
# sample variance of the sample mean
var(population) / length(sample)

```



#$ Law of Large Numbers

As $n$ increases, sample mean approaches population mean.

```{r}
sample_mean_list <- c()
sample_times <- seq(100, 50000, 500)
for (n in sample_times ){
  sample <- sample(population, size = n, replace = TRUE)
  sample_mean_list <- c(sample_mean_list, mean(sample))
}

plot(sample_times, sample_mean_list, xlab = "number of samples", ylab = "sample mean")
abline(h = mean(population), col = "red")
```

#  Central limit theorem
- As $n$ increases, distribution of sample means approaches normal

```{r}

# Now we show the central limit theorem
sample_times <- c(100, 1000, 10000)
col <- c("red", "green", "blue")
i = 1
for (n in sample_times ){
  # central limit theorem talks about distribution of the sample mean
  # we cannot calculate the distribution for a single sample, so we draw sample multiple times
  sample_mean_list <- c()
  for (m in 1:1000)
  {
    sample <- sample(population, size = n, replace = TRUE)
    sample_mean_list <- c(sample_mean_list, mean(sample))
  }
  # sample_mean_list_standard <- sample_mean_list - mean(population)
  plot(density (sample_mean_list),  col = col[i], xlim = c(-0.5, 0.5), ylim  = c(0,50), xlab = "")
  abline(v = mean(population), col = "black")
  par(new = T)
  i = i  + 1
}


```
# bootstrap 

For instance, we care about the *median* of population. Sample median is clearly an estimate of population median. But how can we obtain the *95% confidence interval* of estimated median?
```{r}
sample = sample(population, size = 1000)

sample_median = median(sample)


bootstrap_medians = c() # store bootstrapped medians
for (i in 1:1000){
  # resample from the sample with replacement
  boot_data <- sample(sample, 1000, replace = T)
  boot_median <- median(boot_data)
  bootstrap_medians <- c(bootstrap_medians, boot_median)
}

# then simple quantile function to 
print ("point estimate of median")
print (sample_median)
print ("95% confidence interval of median")
quantile(bootstrap_medians, c(0.025, 0.975))
```

As you may see, the bootstrapped confidence interval may not be symmetric; normal approximated confidence interval, on the other hand, is by definition symmetric.
